---
title: "Import Prov-GigaPath tile level embeddings"
author: "Eslam Abousamra, Ilaria Billato"
date: "2025-11-13"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: ImageFeatureTCGA
vignette: >
  %\VignetteIndexEntry{ImageFeatureTCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r library import, message=FALSE}
library(BiocStyle)
library(ImageFeatureTCGA)
library(ggplot2)
library(dplyr)
library(sfdep)
library(spdep)
library(SpatialExperiment)
library(data.table)
```


# Introduction

In this vignette, we demonstrate how to work with tile-level embeddings derived 
from whole-slide images (WSIs). 
Each tile corresponds to a patch of the tissue, and its embedding is a
high-dimensional vector capturing visual and morphological features 
extracted from Prov-GigaPath.


We begin by importing the tile-level data using ImageFeatureTCGA. 
Next, we perform principal component analysis (PCA) to reduce 
the high-dimensional embeddings to two principal components, 
which facilitates visualization and preliminary exploration of the data. 
We then visualize the spatial layout of the tiles on the tissue slide, 
coloring by the principal components to examine patterns in the embedding space.


# Import Prov-GigaPath tile level embeddings. 

```{r import tile-level emb, results = 'hide'}
# list slides
all_tile_files <- listProvGiga( level = "tile_level")[["Filename"]]

## filter with catalog
getCatalog("provgigapath") |> 
    dplyr::filter(Project.ID == "TCGA-OV") |> 
    dplyr::pull(filename)

# select Ovarian Cancer Slide as an example
tile_prov_url <- paste0(
    "https://store.cancerdatasci.org/provgigapath/tile_level/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.csv.gz"
)

example_slide <- ProvGiga(tile_prov_url) |>
    import()

```

# Embedding PCA
```{r run PCA}

# Extract embedding numbers for pca
embedding_cols <- grep("^[0-9]+$", names(example_slide), value = TRUE)

# Run PCA
pca_res <- prcomp(example_slide[, embedding_cols], scale. = TRUE)

pca_example_slide <- bind_cols(
    example_slide,
    as_tibble(pca_res$x)[, 1:2] |> rename(PC1 = "PC1", PC2 = "PC2")
)

```


```{r visualize PC}
ggplot(pca_example_slide, aes(PC1, PC2)) +
    geom_point(alpha = 0.6, size = 1) +
    theme_minimal() +
    labs(title = "Tile-level PCA Ovarian Cancer Embedding: Single Slide")


ggplot(pca_example_slide, aes(tile_x, tile_y, color = PC1)) +
    geom_point(size = 1) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    labs(title = "Tissue layout colored by PC1")

```

# Spatial Patterns

To investigate spatial patterns in the tissue, we use the PCA-reduced embeddings
for each tile. Each tile has a physical location (tile_x, tile_y) on the slide,
which allows us to explore how similar embedding values cluster across space. 
We construct a k-nearest neighbor graph to define which tiles are spatially 
“connected,” and then compute global and local spatial autocorrelation metrics.


```{r extract coords}
coords <- pca_example_slide[, c("tile_x", "tile_y")]
nb <- knn2nb(knearneigh(coords, k = 6))
lw <- nb2listw(nb, style = "W")

```


Next, we calculate global spatial autocorrelation using Moran’s I and Geary’s C,
which quantify the overall tendency of similar PC1 values to cluster or disperse
on the tissue slide. We also compute Local Moran’s I (LISA) to detect local 
clusters of similar embedding values.

```{r metrics}
mi <- moran.test(pca_example_slide$PC1, lw)
gc <- geary.test(pca_example_slide$PC1, lw)
lisa <- localmoran(pca_example_slide$PC1, lw)
pca_example_slide$localI <- lisa[, "Ii"]
pca_example_slide$localI_pval <- lisa[, "Pr(z != E(Ii))"]

print(mi)
print(gc)
```

We visualize the spatial patterns. The Moran scatterplot shows the relationship 
between each tile’s PC1 value and the mean of its neighbors, while the LISA plot
highlights local clusters (“hotspots”) of high or low PC1 values across the 
tissue slide.

```{r metrics 2}
moran.plot(pca_example_slide$PC1, lw, labels = FALSE,
                main = "Moran scatterplot of PC1")

# LISA visualization
df_lisa <- data.frame(coords, Ii = lisa[, "Ii"])
ggplot(df_lisa, aes(x = tile_x, y = tile_y, color = Ii)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    ggtitle("Local Moran's I (LISA) for PC1")
```



# Adding HoverNet Nuclei Features

You can import HoVerNet segmentation results as a
`SpatialExperiment` or `SpatialFeatureExperiment`.

In this section we want show you hoe to integrate HoVerNet classification and 
segmentation output with Prov-GigaPath embeddings.


```{r importHover}
# import HoVerNet
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/h5ad/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.h5ad.gz"
)

hn_spe <- HoverNet(hov_file, outClass = "SpatialExperiment") |>
    import()

# import Prov-GigaPath
tile_prov_url <- paste0(
    "https://store.cancerdatasci.org/provgigapath/tile_level/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.csv.gz"
)

pg_spe<- ProvGiga(tile_prov_url) |>
    import()


```



```{r cell coordinates}
# Extract cell coordinates from HoVerNet
cell_coords <- spatialCoords(hn_spe)

# Extract nuclei metadata 
cell_meta <- colData(hn_spe)
cell_meta$x <-cell_coords[,1]
cell_meta$y <-cell_coords[,2]
```

# Visualizing Hovernet nuclei vs tile coordinates to see that they do not match 
perfectly. You can use matchHoverNetToTiles to compute the scaling factor.

```{r plotting hovernet vs tiles PCA}
plot(cell_meta$x, cell_meta$y, pch=16, col="#0000FF20")
points(pca_example_slide$tile_x, 
        pca_example_slide$tile_y, 
        pch=16, 
        col="#FF000020")

```


# Scale factor between nuclei coordinates and tile coordinates
```{r compute scaling factor}
match_hv_pg <- matchHoverNetToTiles(hn_spe, pg_spe)

```


```{r visualizing tile level with hovernet}

ggplot(match_hv_pg$tiles_with_nuclei, aes(tile_x, tile_y, 
                                    color = cell_type_label, 
                                    size = N)) +
    geom_point(alpha = 0.7) +
    coord_equal() +
    theme_minimal() +
    labs(title = "All HoverNet cell types per tile")


ggplot(match_hv_pg$tiles_with_nuclei, aes(tile_x, tile_y, 
                                    color = dominant_cell_type)) +
    geom_point(size = 2) +
    coord_equal() +
    theme_minimal() +
    labs(title = "Per-tile dominant HoverNet cell type")


```


# Session Info

```{r sessioninfo}
sessionInfo()
```



